Gammapy: A Python package for gamma-ray astronomy

### 2022-08-08 src/ms-review.tex
Apply suggestions from code review
Co-authored-by: Bruno KHÉLIFI <khelifi@in2p3.fr>
Co-authored-by: Quentin Remy <quentin.remy@live.fr>

%% \gammaray astronomy
\gammaray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us with insight to the
most energetic process in the universe such as surroundings of
black holes and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground based instrument use the Earth's atmosphere as a particle detector.
Very high energy cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on this phenomenon to detect the
primary \gammaray photons and measure their direction and energy.
It covers the energy range from fews tens of GeV up to the PeV.
There are two main categories of instrument \citep{2015CRPhy..16..610D}.
Imaging Atmospheric Cherenkov Telescopes (IACT) make images of atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

@@

%% Context : formats
In practice, the data-reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray like events.
The latter are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the specific instrument response functions (IRF).
The information in this high data level is independent on
the data reduction, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system.

%% Context: develoment of open software
Python has become extremely popular as a scientific programming language
in particular in the field of data sciences. The success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a "glue" language between different programming languages and last but not least
the rich eco-system of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
or optimization packages such as Iminuit \citep{iminuit}.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started in 2015 with the objective of building a flexible and
efficient software library for very high energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a common data format such as \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments the \gadf data format is illustarted in Figure~\ref{fig:big_picture}.

@@

%% Paper outline
In this article, we describe the general structure of the \gammapy package and
its main concepts. In Section \ref{sec:analysis-workflow-overview}, we present
the data analysis workflow in very high energy \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package and
describe the various subpackages it contains. Section \ref{sec:applications}
presents a number of applications, while Section \ref{sec:gammapy-project}
discusses the project organization.

@@

All observation events and instrument responses are projected onto the
user defined geometry. Because residual hadronic background models can be subject
to significant uncertainties, background correction must be applied,
such as the ring or the field-of-view background techniques or
background measurements must be performed within, e.g. reflected regions~\citep{Berge07}.
Parts of the data with high associated IRF systematics must also be excluded by defining
a "safe" data range. These data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see \ref{ssec:gammapy-makers}).

@@

\paragraph{Point spread function:}
\gammapy allows user to treat different kinds of PSFs,
in particular, parametric multi-dimensional gaussians (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing})
one. The \code{EnergyDependentMultiGaussPSF} class is able to handle up to three
gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing takes} into account the gamma and
sigma parameters as defined here. The \code{ParametricPSF} allows to create a
PSF with a representation different from gaussian(s) or King profile(s).
Finally, the user can take advantage from the \code{PSFMap} class, which creates
a multi-dimensional map of the PSF in WCS coordinates. At each position, a PSF
kernel map (\code{PSFKernel}) provides a PSF as a function of the true energy.
The creation of PSF kernel maps, where the PSF is defined for each
sky-position, is also given to the user. The latter two can speed up analyses.

\paragraph{Energy dispersion:}
for \iact, it is generally given in terms of the so-called
 migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. This ratio should be as close as one
and its dispersion can assume the shape of a gaussian (or even more complex
distributions). Migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are \code{EnergyDispersion2D},
designed to interpret Edisp, \code{EDispKernelMap}, which builds an Edisp kernel
map, i.e., a 4-dimensional WCS map where at each sky-position is associated an
Edisp kernel. The latter is a representation of the Edisp as a function of the
true energy only thanks to the sub-class \code{EDispKernel}.

@@

\subsection{gammapy.maps}
\label{ssec:gammapy-maps}
The \code{gammapy.maps} sub-package provides classes that represent data
structures associated with a set of coordinates or a region on a sphere. In
addition it allows to handle an arbitrary number of non-spatial data
dimensions, such as time or energy. It is organized around three types of
structures: geometries, sky-maps and map axes, which inherit from the base
classes \code{Geom}, \code{Map} and \code{MapAxis} respectively.

@@

\subsubsection{HEALPix Maps}
This pixelization scheme ~\citep{Calabretta2002} provides a
subdivision of a sphere in which each pixel covers the same surface area as
every other pixel. As a consequence, however, pixel shapes are no longer
rectangular, or regular.
This pixelisation is typically used for all-sky data, such as data
from the \hawc or \fermi observatory. \gammapy natively supports
the multiscale definiton of the HEALPix pixelisation and thus
allows for easy up and downsampling of the data. In addition to
the all-sky map, \gammapy also supports a local HEALPix
pixelisation where the size of the map is constrained to a given
radius.
For local nighbourhood operations, such as convolution \gammapy relies
on projecting the HEALPix data to a local tangential WCS grid.

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow access by name later in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The end product of the data reduction process described in Section
\ref{ssec:gammapy-makers} are a set of binned counts, background and IRF maps,
at the DL4 level. The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with associated models and the likelihood, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistics, different types of Datasets are supported. \code{MapDataset} is
used for 3D (spectral and morphological) fitting, and a 1D spectral fitting is
done using \code{SpectrumDataset}. While the default statistics for both of these is
Cash, their corresponding on off versions are adapted for the case where the
background is measured from real off counts, and support wstat statistics. The
predicted counts are computed by convolution of the models with the associated
IRFs. Fitting of precomputed flux points is enabled through \code{FluxPointsDataset},
using chi2 statistics. Multiple datasets of same or different types can be
bundled together in \code{Datasets} (e.g., Figure \ref{fig*:minted:gp_datasets}),
where the likelihood from each constituent member is added, thus facilitating
joint fitting across different observations, and even different instruments
across different wavelengths. Datasets also provide functionalities for
manipulating reduced data, e.g. stacking, sub-grouping, plotting. Users can
also create their customized datasets for implementing modified likelihood
methods.

@@

The model gallery provides a visual overview of the available models in
\gammapy. Most of the analytic models  commonly used in \gammaray astronomy are
built-in. We also offer a wrapper to radiative models implemented in the Naima
package~\cite{naima}. The modeling framework can be easily extended with
user-defined models. For example agnpy models that describe leptonic radiative
processes in jetted Active Galactic Nuclei (AGN) can wrapped into
\gammapy~\citep[see Section 3.5 of ][]{2021arXiv211214573N} .

@@

The \code{Fit} class provides methods to fit i.e., optimise parameters and estimate
parameter errors and correlations. It interfaces with a Datasets object, which
in turn is connected to a Models object containing the model parameters in its
Parameters object. Models can be unique for a given dataset, or contribute to
multiple datasets and thus provide links, allowing e.g., to do a joint fit to
multiple IACT datasets, or to a joint IACT and \fermi dataset. Many
examples are given in the tutorials.

@@

\subsection{gammapy.estimators}
\label{ssec:gammapy-estimators}
By fitting parametric models to the data, the total integrated
flux and overall temporal, spectral and morphological shape of the
\gammaray emission can be constrained. In many cases it useful
to make a more detailed follow-up analysis by measuring the
flux in smaller spectral, temporal or spatial bins. This
possibly reveals more detailed emission features, which
are relevant for studying correlation with counterpart emissions.

The \code{gammapy.estimators} sub-module features methods to compute flux
points, light curves, flux maps, flux profiles from data.
The basic method for all these measurements is equivalent.
The initial fine bins of \code{MapDataset} are grouped into
larger bins. The norm of the best fit "reference" spectral
model is fitted ("scaled") in the  restricted data range only.

@@

Optionally it can also compute more advanced quantities
such as assymmetric flux errors, flux upper limits
and one dimensional profiles of the fit statistic,
which show how the likelihood functions varies with
the flux norm parameter around the fit minimum.
This information is useful for inspecting the quality
of the fit, where assumptoticcally a parabolic
shape of the profile is expected at the best fit
values (\todo{Reference to modeling / fitting section?}).

The result of the flux point estimation are either stored in a
\code{FluxMaps} or \code{FluxPoints} object. Both objects
are based on an internal representation of the flux which is
independent of the SED type. The flux is represented by a
the reference spectral model and an array of
normalisation values given in energy, time and spatial bins,
which factorise the deviation of the flux in the given
bin form the reference spectral model. This allows
user to conveniently transform between different
SED types. Table~\ref{tab:sed_types} shows an
overview and definitions of the supported SED types.
The actual flux values for each SED type are obtained
by mutiplication of the norm with the  reference flux.

\begin{table*}
    \begin{center}
        \begin{tabular}{lll}
         \hline
         Type & Description & Unit Equivalency \\
         \hline
         dnde & Differential flux at a given energy & $\mathrm{TeV^{-1}~cm^{2}~s^{-1}}$ \\
         e2dnde & Differential flux at a given energy  & $\mathrm{TeV~cm^{2}~s^{-1}}$ \\
         flux & Integrated flux in agiven energy range & $\mathrm{cm^{2}~s^{-1}}$ \\
         eflux & Integrated energy flux in agiven energy range & $\mathrm{erg~cm^{2}~s^{-1}}$\\
         \hline
        \end{tabular}
    \end{center}
    \label{tab:sed_types}
    \caption{Definition of the different SED types supported in \gammapy.}
\end{table*}

Both result objects support the possibility to serialise
the data into multiple formats. This includes the
\gadf SED (reference?) format, FITS based ND sky-maps
and formats compatible with Astropy's \code{Table} and
\code{BinnedTimeSeries} data structures. This allows
users to directly further analyse the results, e.g.
standard algorithms for time analysis such as
computing Lomb-Scargle periodograms or Bayesian
blocks for time series. So far \gammapy does not
support "unfolding" of \gammaray spectra.
Methods for this will be implemented in future version of \gammapy.

@@

\subsection{gammapy.analysis}
\label{ssec:gammapy-analysis}
The \code{gammapy.analysis} sub-module provides a  high-level interface for the most
common use cases identified in \gammaray analyses. The classes and methods
included can be used in Python scripts, notebooks or as commands within IPython
sessions. The high-level user interface can also be used to automatise
workflows driven by parameters declared in a configuration file in YAML format.
This way a full analysis can be executed via a single command line taking the
configuration file as input.

The \code{Analysis} class has the responsibility of orchestrating of the workflow
defined in the configuration \code{AnalysisConfig} objects and triggering the execution of
the \code{AnalysisStep} classes that defines the common use cases identified. These
steps include the following: observations selection with the \code{DataStore},  data
reduction, excess map computation, model fitting, flux points estimation, and
light curves production. The structure of the \code{Analysis} class allows to define
custom AnalysisStep, so one could design and configure arbitrarily complex
workflows. \todo{the latter is not true for v1.0..., include example YAML file?}

@@

The sub-package also provides matplotlib implementations of specific
colormaps for false color image representation. Those colormaps have
been used historically larger collaborations in the very high energy
domain (such as \milagro or \hess) as "trademark" colormaps.
While we explicitly discourage the use of those colormaps for publication
of new results, because they do not follow modern visualization
standards, such as linear brightness gradients and accessibility
for visually impaired people, we still consider the colormaps
useful for reproducibility of past results.

\subsection{gammapy.astro}
\label{ssec:gammapy-astro}
The \code{gammapy.astro} sub-package contains utility functions for studying physical
scenarios in high energy astrophysics. The \code{gammapy.astro.darkmatter} module
computes the so  called J-factors and the associated \gammaray spectra expected
from annihilation of dark matter in different channels according to the recipe
described in \cite{2011JCAP...03..051C}.

In the \code{gammapy.astro.source} sub-module, dedicated classes exist for modeling
galactic \gammaray sources according to simplified physical models, eg: SNR evolution
models \citep{1950RSPSA.201..159T, 1999ApJS..120..299T}, evolution of PWN during the
free expansion phase \citep{2006ARA&A..44...17G} or computation
of physical parameters in a pulsar assumed to be a simple dipole.

In the \code{gammapy.astro.population} sub-module there are dedicated tools
for simulating synthetic populations based on physical models derived from
observational or theoretical considerations for different classes of Galactic
VHE \gammaray emitters: PWNe, SNRs \cite{1998ApJ...504..761C},
pulsars \cite{2006ApJ...643..332F, 2006MNRAS.372..777L, 2004A&A...422..545Y}
and \gammaray binaries.

@@

\subsection{gammapy.catalog}
\label{ssec:gammapy-catalog}
Comprehensive source catalogs are increasingly being provided by many high
energy astrophysics experiments. The \code{gammapy.catalog} sub-packages
provides a convenient access to the most important \gammaray catalogs.
Catalogs are represented by the \code{SourceCatalog} object, which
contains the actual catalog as an Astropy \code{Table} object.
Objects in the catalog can be accesed by row index, name of the
object or any association or alias name listed in the catalog.

Sources are represented in \gammapy by the \code{SourceCatalogObject}
class, which has the responsibility to translate the information
contained in the catalog to other \gammapy objects. This includes
the spatial and spectral model of the source, flux points and
light curves (if available) for individual objects. This
module works independently from the rest of the package, and the required
catalogs are supplied in \code{GAMMAPY\_DATA} repository.
The overview of currenly supported catalogs, the corresponding
\gammapy classes and references are shown in Table~\ref{tab:catalogs}.
Newly released relevant catalogs will be added in future.

\begin{table*}[ht!]
    \begin{center}
        \begin{tabular}{llll}
         \hline
         Class Name & Shortcut & Description & Reference\\
         \hline
         \code{SourceCatalog3FGL} & \code{"3fgl"} & 3\textsuperscript{rd} catalog of \fermi sources & \cite{3FGL} \\
         \code{SourceCatalog4FGL} & \code{"4fgl"} & 4\textsuperscript{th} catalog of \fermi  sources & \cite{4FGL} \\
         \code{SourceCatalog2FHL} & \code{"2fhl"} & 2\textsuperscript{nd} catalog high energy \fermi  sources & \cite{2FHL} \\
         \code{SourceCatalog3FHL} & \code{"3fhl"} & 3\textsuperscript{rd} catalog high energy \fermi  sources & \cite{3FHL} \\
         \code{SourceCatalog2HWC} & \code{"2hwc"} & 2\textsuperscript{nd} catalog of \hawc sources & \cite{2HWC} \\
         \code{SourceCatalog3HWC} & \code{"3hwc"} & 3\textsuperscript{rd} catalog of \hawc sources & \cite{3HWC} \\
         \code{SourceCatalogHGPS} & \code{"hgps"} & \hess Galactic Plane Survey catalog & \cite{HGPS} \\
         \code{SourceCatalogGammaCat} & \code{"gammacat"} & Open source data collection & \citep{gamma-cat} \\
         \hline
         \end{tabular}
    \end{center}
    \label{tab:catalogs}
    \caption{Overview of supported catalogs in \code{gammapy.catalog}.}
\end{table*}

@@

For such situations a more complex approach is needed, the so-called 3D
analysis. The three relevant dimensions are the two spatial angular coordinates
and an energy axis. In this framework, a combined spatial and spectral model
(that is, a \code{SkyModel}, see Section~\ref{ssec:gammapy-modeling}) is fitted to the
sky-maps that were previously derived from the data and bundled into a
\code{MapDataset} (see Sections~\ref{ssec:gammapy-makers} and~\ref{ssec:gammapy-datasets}).

>>>>

%% \gammaray astronomy
\gammaray astronomy is a rather young field of research. The \gammaray~
range of the electromagnetic spectrum provides us with insight to the
most energetic process in the universe such as surroundings of
black holes and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground based instrument use the Earth's atmosphere as a particle detector.
Very high-energy cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on this phenomenon to detect the
primary \gammaray photons and measure their direction and energy.
It covers the energy range from fews tens of GeV up to the PeV.
There are two main categories of instrument \citep{2015CRPhy..16..610D}.
Imaging Atmospheric Cherenkov Telescopes (IACT) make images of atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

@@

%% Context : formats
In practice, the data reduction workflow of all \gammaray observatories
is remarkably similar. After data calibration, shower events are reconstructed and
gamma/hadron separation is applied to build lists of \gammaray like events.
The latter are then used to derive scientific results, such as spectra, sky maps
or light curves, taking into account the specific instrument response functions (IRF).
The information in this high data level is independent on
the data reduction, and eventually of the detection technique. This implies,
for example, that data from IACT and particle samplers can be represented
within the same model. The efforts to prototype a format usable by various instruments
converged in the so-called \textit{Data Format for \gammaray Astronomy}
initiative~\citep{gadf_proc, gadf_universe}, abbreviated in
\texttt{gamma-astro-data-format} (\gadf). The latter proposes prototypical
specifications to produce files based on the flexible image transport system
(FITS) format~\citep{fits} encapsulating this high-level information. This is
realized by storing a list of gamma-like events with their measured quantities
(energy, direction, arrival time) and a parametrisation of the response of the
system.

%% Context: develoment of open software
Python has become extremely popular as a scientific programming language
in particular in the field of data sciences. The success is
mostly attributed to the simple and easy to learn syntax, the ability to act as
a 'glue' language between different programming languages and last but not least
the rich eco-system of packages and its open and supportive community, from
core numerical analysis packages such as \numpy~\citep{numpy} and
\scipy~\citep{2020SciPy-NMeth}, visualization libraries, e.g. Matplotlib \citep{matplotlib},
or optimization packages such as Iminuit \citep{iminuit}.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project started in 2015 with the objective of building a flexible and
efficient software library for very high-energy \gammaray data analysis \citep{gammapy_2015}.
This is possible thanks to the definition of a common data format such as \gadf, and thanks to
an open development with a broad community of contributors. The relation of \gammapy to the
different instruments the \gadf data format is illustarted in Figure~\ref{fig:big_picture}.

@@

%% Paper outline
In this article, we describe the general structure of the \gammapy package and
its main concepts. In Section \ref{sec:analysis-workflow-overview}, we present
the data analysis workflow in very high-energy \gammaray astronomy. Then we
show how this workflow is reflected in the structure of the \gammapy package and
describe the various subpackages it contains. Section \ref{sec:applications}
presents a number of applications, while Section \ref{sec:gammapy-project}
discusses the project organization.

@@

All observation events and instrument responses are projected onto the
user defined geometry. Because residual hadronic background models can be subject
to significant uncertainties, background correction must be applied,
such as the ring or the field-of-view background techniques or
background measurements must be performed within, e.g. reflected regions~\citep{Berge07}.
Parts of the data with high associated IRF systematics must also be excluded by defining
a 'safe' data range. These data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see \ref{ssec:gammapy-makers}).

@@

\paragraph{Point spread function:}
\gammapy allows user to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussians (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing})
one. The \code{EnergyDependentMultiGaussPSF} class is able to handle up to three
Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing takes} into account the gamma and
sigma parameters as defined here. The \code{ParametricPSF} allows to create a
PSF with a representation different from Gaussian(s) or King profile(s).
Finally, the user can take advantage from the \code{PSFMap} class, which creates
a multi-dimensional map of the PSF in WCS coordinates. At each position, a PSF
kernel map (\code{PSFKernel}) provides a PSF as a function of the true energy.
The creation of PSF kernel maps, where the PSF is defined for each
sky-position, is also given to the user. The latter two can speed up analyses.

\paragraph{Energy dispersion:}
for \iact, it is generally given in terms of the so-called
 migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. This ratio should be as close as one
and its dispersion can assume the shape of a Gaussian (or even more complex
distributions). Migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are \code{EnergyDispersion2D},
designed to interpret Edisp, \code{EDispKernelMap}, which builds an Edisp kernel
map, i.e., a 4-dimensional WCS map where at each sky-position is associated an
Edisp kernel. The latter is a representation of the Edisp as a function of the
true energy only thanks to the sub-class \code{EDispKernel}.

@@

\subsection{gammapy.maps}
\label{ssec:gammapy-maps}
The \code{gammapy.maps} sub-package provides classes that represent data
structures associated with a set of coordinates or a region on a sphere. In
addition it allows to handle an arbitrary number of non-spatial data
dimensions, such as time or energy. It is organized around three types of
structures: geometries, sky maps and map axes, which inherit from the base
classes \code{Geom}, \code{Map} and \code{MapAxis} respectively.

@@

\subsubsection{HEALPix Maps}
This pixelization scheme ~\citep{Calabretta2002} provides a
subdivision of a sphere in which each pixel covers the same surface area as
every other pixel. As a consequence, however, pixel shapes are no longer
rectangular, or regular.
This pixelisation is typically used for all-sky data, such as data
from the \hawc or \fermi observatory. \gammapy natively supports
the multiscale definiton of the HEALPix pixelisation and thus
allows for easy up and downsampling of the data. In addition to
the all-sky map, \gammapy also supports a local HEALPix
pixelisation where the size of the map is constrained to a given
radius.
For local neighbourhood operations, such as convolution \gammapy relies
on projecting the HEALPix data to a local tangential WCS grid.

@@

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow access by name later in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The end product of the data reduction process described in Section
\ref{sec:gammaray-data-analysis} (and in detail in \ref{ssec:gammapy-makers}) are a set of binned counts, background and IRF maps,
at the DL4 level. The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with associated models and the likelihood, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistics, different types of Datasets are supported. \code{MapDataset} is
used for 3D (spectral and morphological) fitting, and a 1D spectral fitting is
done using \code{SpectrumDataset}. While the default statistics for both of these is
Cash, their corresponding on off versions are adapted for the case where the
background is measured from real off counts, and support wstat statistics. The
predicted counts are computed by convolution of the models with the associated
IRFs. Fitting of precomputed flux points is enabled through \code{FluxPointsDataset},
using chi2 statistics. Multiple datasets of same or different types can be
bundled together in \code{Datasets} (e.g., Figure \ref{fig*:minted:gp_datasets}),
where the likelihood from each constituent member is added, thus facilitating
joint fitting across different observations, and even different instruments
across different wavelengths. Datasets also provide functionalities for
manipulating reduced data, e.g. stacking, sub-grouping, plotting. Users can
also create their customized datasets for implementing modified likelihood
methods.

@@

The model gallery provides a visual overview of the available models in
\gammapy. Most of the analytic models  commonly used in \gammaray astronomy are
built-in. We also offer a wrapper to radiative models implemented in the Naima
package~\cite{naima}. The modeling framework can be easily extended with
user-defined models. For example agnpy models that describe leptonic radiative
processes in jetted Active Galactic Nuclei (AGN) can be wrapped into
\gammapy~\citep[see Section 3.5 of ][]{2021arXiv211214573N} .

@@

The \code{Fit} class provides methods to fit i.e., optimise parameters and estimate
parameter errors and correlations. It interfaces with a \code{Datasets} object, which
in turn is connected to a \code{Models} object containing the model parameters in its
\code{Parameters} object. Models can be unique for a given dataset, or contribute to
multiple datasets and thus provide links, allowing e.g., to do a joint fit to
multiple IACT datasets, or to a joint IACT and \fermi dataset. Many
examples are given in the tutorials.

@@

\subsection{gammapy.estimators}
\label{ssec:gammapy-estimators}
By fitting parametric models to the data, the total integrated
flux and overall temporal, spectral and morphological shape of the
\gammaray emission can be constrained. In many cases it is useful
to make a more detailed follow-up analysis by measuring the
flux in smaller spectral, temporal or spatial bins. This
possibly reveals more detailed emission features, which
are relevant for studying correlation with counterpart emissions.

The \code{gammapy.estimators} sub-module features methods to compute flux
points, light curves, flux maps, flux profiles from data.
The basic method for all these measurements is equivalent.
The initial fine bins of \code{MapDataset} are grouped into
larger bins. The norm of the best fit "reference" spectral
model is fitted (or 'scaled') in the  restricted data range only.

@@

Optionally it can also compute more advanced quantities
such as assymmetric flux errors, flux upper limits
and one dimensional profiles of the fit statistic,
which show how the likelihood functions varies with
the flux norm parameter around the fit minimum.
This information is useful for inspecting the quality
of a fit, for which asymptomatically a parabolic
shape of the profile is expected at the best fit
values (\todo{Reference to modeling / fitting section?}).

The result of the flux point estimation are either stored in a
\code{FluxMaps} or \code{FluxPoints} object. Both objects
are based on an internal representation of the flux which is
independent of the Spectral Energy Distribution (SED) type. The flux is represented by a
the reference spectral model and an array of
normalisation values given in energy, time and spatial bins,
which factorises the deviation of the flux in a given
bin from the reference spectral model. This allows
user to conveniently transform between different
SED types. Table~\ref{tab:sed_types} shows an
overview and definitions of the supported SED types.
The actual flux values for each SED type are obtained
by mutiplication of the norm with the  reference flux.

\begin{table*}
    \begin{center}
        \begin{tabular}{lll}
         \hline
         Type & Description & Unit Equivalency \\
         \hline
         dnde & Differential flux at a given energy & $\mathrm{TeV^{-1}~cm^{-2}~s^{-1}}$ \\
         e2dnde & Differential flux at a given energy  & $\mathrm{TeV~cm^{-2}~s^{-1}}$ \\
         flux & Integrated flux in a given energy range & $\mathrm{cm^{-2}~s^{-1}}$ \\
         eflux & Integrated energy flux in a given energy range & $\mathrm{erg~cm^{-2}~s^{-1}}$\\
         \hline
        \end{tabular}
    \end{center}
    \label{tab:sed_types}
    \caption{Definition of the different SED types supported in \gammapy.}
\end{table*}

Both result objects support the possibility to serialise
the data into multiple formats. This includes the
\gadf SED (reference?) format, FITS based ND sky-maps
and formats compatible with Astropy's \code{Table} and
\code{BinnedTimeSeries} data structures. This allows
users to directly further analyse the results, e.g.
standard algorithms for time analysis such as
computing Lomb-Scargle periodograms or Bayesian
blocks for time series. So far \gammapy does not
support 'unfolding' of \gammaray spectra.
Methods for this will be implemented in a future version of \gammapy.

@@

\subsection{gammapy.analysis}
\label{ssec:gammapy-analysis}
The \code{gammapy.analysis} sub-module provides a high-level interface (HLI) for the most
common use cases identified in \gammaray analyses. The included classes and methods
 can be used in Python scripts, notebooks or as commands within IPython
sessions. The HLI can also be used to automatise
workflows driven by parameters declared in a configuration file in YAML format.
This way, a full analysis can be executed via a single command line taking the
configuration file as input.

The \code{Analysis} class has the responsibility of orchestrating of the workflow
defined in the configuration \code{AnalysisConfig} objects and triggering the execution of
the \code{AnalysisStep} classes that define the identified common use cases. These
steps include the following: observations selection with the \code{DataStore},  data
reduction, excess map computation, model fitting, flux points estimation, and
light curves production. The structure of the \code{Analysis} class allows to define
custom AnalysisStep, so one could design and configure arbitrarily complex
workflows. \todo{the latter is not true for v1.0..., include example YAML file?}

@@

The sub-package also provides "matplotlib" implementations of specific
colormaps for false color image representation. Those colormaps have
been used historically by larger collaborations in the very high-energy
domain (such as \milagro or \hess) as 'trademark' colormaps.
While we explicitly discourage the use of those colormaps for publication
of new results, because they do not follow modern visualization
standards, such as linear brightness gradients and accessibility
for visually impaired people, we still consider the colormaps
useful for reproducibility of past results.

\subsection{gammapy.astro}
\label{ssec:gammapy-astro}
The \code{gammapy.astro} sub-package contains utility functions for studying physical
scenarios in high-energy astrophysics. The \code{gammapy.astro.darkmatter} module
computes the so called J-factors and the associated \gammaray spectra expected
from annihilation of dark matter in different channels according to the recipe
described in \cite{2011JCAP...03..051C}.

In the \code{gammapy.astro.source} sub-module, dedicated classes exist for modeling
galactic \gammaray sources according to simplified physical models, e.g. SuperNova Remnant (SNR) evolution
models \citep{1950RSPSA.201..159T, 1999ApJS..120..299T}, evolution of Pulsar Wind Nebula (PWN) during the
free expansion phase \citep{2006ARA&A..44...17G} or computation
of physical parameters in a pulsar assumed to be a simple dipole.

In the \code{gammapy.astro.population} sub-module there are dedicated tools
for simulating synthetic populations based on physical models derived from
observational or theoretical considerations for different classes of Galactic
very high-energy \gammaray emitters: PWNe, SNRs \cite{1998ApJ...504..761C},
pulsars \cite{2006ApJ...643..332F, 2006MNRAS.372..777L, 2004A&A...422..545Y}
and \gammaray binaries.

@@

\subsection{gammapy.catalog}
\label{ssec:gammapy-catalog}
Comprehensive source catalogs are increasingly being provided by many high-energy
 astrophysics experiments. The \code{gammapy.catalog} sub-packages
provides a convenient access to the most important \gammaray catalogs.
Catalogs are represented by the \code{SourceCatalog} object, which
contains the actual catalog as an Astropy \code{Table} object.
Objects in the catalog can be accesed by row index, name of the
object or any association or alias name listed in the catalog.

Sources are represented in \gammapy by the \code{SourceCatalogObject}
class, which has the responsibility to translate the information
contained in the catalog to other \gammapy objects. This includes
the spatial and spectral models of the source, flux points and
light curves (if available) for each individual object. This
module works independently from the rest of the package, and the required
catalogs are supplied in \code{GAMMAPY\_DATA} repository.
The overview of currenly supported catalogs, the corresponding
\gammapy classes and references are shown in Table~\ref{tab:catalogs}.
Newly released relevant catalogs will be added in future.

\begin{table*}[ht!]
    \begin{center}
        \begin{tabular}{llll}
         \hline
         Class Name & Shortcut & Description & Reference\\
         \hline
         \code{SourceCatalog3FGL} & \code{"3fgl"} & 3\textsuperscript{rd} catalog of \fermi sources & \cite{3FGL} \\
         \code{SourceCatalog4FGL} & \code{"4fgl"} & 4\textsuperscript{th} catalog of \fermi  sources & \cite{4FGL} \\
         \code{SourceCatalog2FHL} & \code{"2fhl"} & 2\textsuperscript{nd} catalog high-energy \fermi  sources & \cite{2FHL} \\
         \code{SourceCatalog3FHL} & \code{"3fhl"} & 3\textsuperscript{rd} catalog high-energy \fermi  sources & \cite{3FHL} \\
         \code{SourceCatalog2HWC} & \code{"2hwc"} & 2\textsuperscript{nd} catalog of \hawc sources & \cite{2HWC} \\
         \code{SourceCatalog3HWC} & \code{"3hwc"} & 3\textsuperscript{rd} catalog of \hawc sources & \cite{3HWC} \\
         \code{SourceCatalogHGPS} & \code{"hgps"} & \hess Galactic Plane Survey catalog & \cite{HGPS} \\
         \code{SourceCatalogGammaCat} & \code{"gammacat"} & Open source data collection & \citep{gamma-cat} \\
         \hline
         \end{tabular}
    \end{center}
    \label{tab:catalogs}
    \caption{Overview of supported catalogs in \code{gammapy.catalog}.}
\end{table*}

@@

For such situations a more complex approach is needed, the so-called 3D
analysis. The three relevant dimensions are the two spatial angular coordinates
and an energy axis. In this framework, a combined spatial and spectral model
(that is, a \code{SkyModel}, see Section~\ref{ssec:gammapy-modeling}) is fitted to the
sky maps that were previously derived from the data reduction step and bundled into a
\code{MapDataset} (see Sections~\ref{ssec:gammapy-makers} and~\ref{ssec:gammapy-datasets}).
